### 2014-10-19 Introduction.tex
Pass through Introduction

The goal of this work is to provide a framework for full verification of such
complex hardware systems. Hardware designers already recognize the challenge of
implementing such systems, to such an extent that they have already become some
of the most serious real-world adopters of formal methods.  Hardware
verification is dominated by model-checking, where abstraction techniques are
used to reduce designs to finite state spaces, which can be explored
exhaustively.  There are limits to the construction of sound abstractions, so
verifications of, e.g., cache-coherence protocols have almost always treated
systems with concrete topologies, involving particular finite numbers of caches
and processors. For instance, explicit-state model checking tools like
Murphi~\cite{murphi} or TLC~\cite{tlc} are only able to handle small and
simplified systems, \eg{} a single-level cache hierarchy with three CPUs, two
addresses, and one bit of data. Symbolic model checking techniques have fared
slightly better; McMillan \etal{} have verified a 2-level MSI protocol based on
the Gigamax distributed multiprocessor using SMV~\cite{gigamax}. Optimizations
on these techniques (\eg{} partial order reduction~\cite{part}, symmetry
reduction~\cite{sym1, sym2}, compositional reasoning~\cite{comp}, \etc{}) scale
the approach further, but they are still unable to handle realistic
hierarchical protocols. We might hope to achieve higher assurance and
understanding of design ideas by verifying \emph{infinite families} of hardware
designs, which resist reduction to finite-state models.  A potential related
benefit is reducing the computational resources required to formally verify a
new design in the family, applying a more general theorem without much new
analysis.

Modularity has long been understood as a key property for effective design and
verification of such complex systems. For the designer, it allows a separation
of concerns increasing robustness by allowing the behavior encapsulated by a
modular boundary to be realized by multiple implementation any of which may be
safely dropped into the system. It also allows a greater measure of
parallelization of design improving development time. Similarly verification is
simplified as modular interface agreements provide a set of natural lemmas to
verify. This leads us towards a decomposition of the whole system verification
task into lemmas of sub-systems, which can be composed in a black-box manner to
produce full-system theorems.

This works introduces an interface enabling true modular implementation and
verification of shared memory systems. We want our modular system to permit
\emph{black-box refinement}, that is, nothing beyond the interface contract can
be assumed about other modules.  That is, we should be able to describe a
modular decomposition where we may replace either the processor nodes, the
memory with locally verified alternatives without changing the correctness of
our whole system.  To this end, we identify a key property called \emph{store
atomicity} shared by the memory component of every memory model. This property
effectively restricts the behavior of the memory sub-system to that of an
\emph{atomic instantaneous memory}. We prove that this property is in fact what
is guaranteed using the cache coherence protocols in a multi-processor system
containing a hierarchy of caches, enabling replacement of the cache hierarchy
with an atomic instantaneous memory. Similarly, on the processor side, we
define what is meant to be a correct single processor. These invariants about
the processor and the memory sub-system remains constant across memory models.
We show a concrete instance of this framework by proving that a highly
speculative multi-processor system connected to a coherent cache hierarchy
implements the sequential consistency memory model.

Our proof methodology involves describing a reference model of a component and
an implementation model for that component using labeled transition systems
(LTS) and formally proving that the implementation is correct (sound) with
respect to our reference. Using LTS for describing hardware has already been
studied \cite{HoeArvind:TRSSynthesis1, Hoe:TCAD}, and there are compilers converting an LTS into efficient
hardware \cite{BSV:LangRef, Bluespec:TFRG}. In fact, a cache coherence protocol implementation in Bluespec was discussed in \cite{DNA:CoherenceImplementation}. In fact, the cache coherence protocol that we
study in this work is obtained from a direct transliteration of Bluespec code
used to implement a multi-processor PowerPC system \cite{Khan:PowerPc}. The
resulting implementation from that Bluespec code is pretty efficient; an 8-core
system can run 55 million instructions overall per second on the BEE FPGA
board, with a 2-level cache hierarchy. All our proofs are mechanized in the Coq
proof assistant, and are included as the non-anonymous supplement to this
paper.

@@

\item A general \emph{modular verification methodology} applicable to hardware
  algorithms, based on \emph{labeled transition systems}.  We are able to verify
  the several components of Figure~\ref{zoom} against general formal interfaces,
  enabling mixing and matching of different realizations of each interface,
  without doing any new proofs that peek beneath abstraction boundaries.

@@

\paragraph{Paper Organization:} In Section~\ref{sec:lts} we introduce our
flavor of the labeled transition systems formalism, including a definition of
trace refinement. In Section~\ref{sec:store-atomicity}, we show a generic
decomposition of any multi-processor system independent of the memory model
that it implements, and discuss the store-atomicity property of the memory
sub-component. In Section~\ref{sec:sc} we
give a simple formal model of sequential consistency.  The
following sections refine the two main subcomponents from Figure~\ref{zoom}.
Section~\ref{sec:ooo} discusses definition and verification of a speculative
processor model, and Sections~\ref{sec:cc} and~\ref{sec:ccproof} respectively
define and prove our hierarchical cache-coherence protocol.  Finally, in
Section~\ref{sec:finalresult} we show the whole-system modular proof of our
complex system and a discussion of related work and other conclusions in
Sections~\ref{relatedWork} and \ref{sec:conclusion}.

>>>>

The goal of this work is to provide a framework for full verification of such
complex hardware systems. Hardware designers already recognize the challenge of
implementing these systems, to such an extent that they have already become some
of the most serious real-world adopters of formal methods.  Hardware
verification is dominated by model-checking, where abstraction techniques are
used to reduce designs to finite state spaces, which can be explored
exhaustively.  There are limits to the construction of sound abstractions, so
verifications of, e.g., cache-coherence protocols have almost always treated
systems with concrete topologies, involving particular finite numbers of caches
and processors. For instance, explicit-state model checking tools like
Murphi~\cite{murphi} or TLC~\cite{tlc} are only able to handle small and
simplified systems, \eg{} a single-level cache hierarchy with three CPUs, two
addresses, and one bit of data. Symbolic model checking techniques have fared
slightly better: McMillan \etal{} have verified a 2-level MSI protocol based on
the Gigamax distributed multiprocessor using SMV~\cite{gigamax}. Optimizations
on these techniques (\eg{} partial order reduction~\cite{part}, symmetry
reduction~\cite{sym1, sym2}, compositional reasoning~\cite{comp}, \etc{}) scale
the approach further, but they are still unable to handle realistic
hierarchical protocols. We might hope to achieve higher assurance and
understanding of design ideas by verifying \emph{infinite families} of hardware
designs, which resist reduction to finite-state models.  A potential related
benefit is reducing the computational resources required to formally verify a
new design in the family, applying a more general theorem without much new
analysis.

Modularity has long been understood as a key property for effective design and
verification of such complex systems. For the designer, it allows a separation
of concerns, increasing robustness by allowing the behavior encapsulated by a
modular boundary to be realized by multiple implementations, any of which may be
dropped into the system safely. It also allows a greater measure of
parallelization of human design effort, improving development time. Similarly, verification is
simplified, as modular interface agreements provide a natural lemma structure.
This leads us towards a decomposition of the whole system verification
task into lemmas of subsystems, which can be composed in a black-box manner to
produce full-system theorems.

In this paper, we introduce \textbf{a new approach to modular,
  deductive verification of hardware designs} at the highest level of
rigor, backed by proofs using the Coq proof assistant.  As a
challenging case study, we focus on \textbf{verifying that a
  particular infinite family of multicore, shared-memory systems
  implements sequential consistency}.  The proof is parametrized over
an unknown number of processors connected to an arbitrary memory
hierarchy with an unknown number of caches in an unknown number of
layers (e.g., L1, L2).  In our design, processors and memory systems
both independently employ intricate optimizations that exploit
opportunities for parallelism.  We are able to prove that each of
those two main components still implements strong enough semantics to
support SC, and then we compose those theorems into a result for the
full system.  Either component may be optimized further without
requiring any changes to the implementation, specification, or proof
of the other.

To structure our hardware verification framework, we import
\emph{labeled transition systems (LTSes)}, a well-studied approach from the
world of concurrent software modeling.  As our unified notion of
specification and proof, we adopt \emph{trace refinement}, which
captures when one concurrent system can only produce observable IO
behaviors that another system could also produce.  Each of our
realistic hardware components is associated with a simpler ``reference
implementation'' that serves as a specification, and we prove that
each realistic component refines its spec.  Thereafter, it is sound to
use the simpler spec component in reasoning about the behavior of the
system.

Each key component of our case-study system has a simple spec: the
ideal memory responds instantly and atomically to each load or store
request, and the ideal processor executes instructions in order with
no speculation.  The optimized implementation of each is much more
complicated: the cache-hierarchy memory system potentially allows
every cache to be busy simultaneously computing and sending messages,
and the optimized speculative processor sends multiple load requests
to the memory at once, without waiting for responses first.  The meat
of the verification is in showing that each optimized component
refines its simpler counterpart.  After each such proof, we may in
effect substitute the optimized version for the simple version in a
black-box way within a design, with a full guarantee of soundness.

LTSes as hardware descriptions are an established
idea~\cite{HoeArvind:TRSSynthesis1, Hoe:TCAD}, and there are
compilers that convert LTSes into efficient hardware.  Our work is
based on the Bluespec language~\cite{BSV:LangRef, Bluespec:TFRG},
which served as the model for the formalism of this paper.  Bluespec
models hardware components as atomic rules of a transition system over
state elements, and there is a commercial compiler from such code into
register transfer-level circuits (i.e., Verilog code) with competitive
performance.  Our cache-coherent memory system is directly
transliterated from a Bluespec
implementation~\cite{DNA:CoherenceImplementation} used to
implement a multi-processor PowerPC system~\cite{Khan:PowerPc}. The
hardware synthesized from that implementation is rather efficient:
an 8-core system can run 55 million instructions per second on
the BEE FPGA board\cite{}, with a 2-level cache hierarchy.

@@

\item A general \emph{modular verification methodology} applicable to hardware
  algorithms, based on \emph{labeled transition systems}.  We are able to verify
  the several components of Figure~\ref{zoom} \adam{OK to have a
    forward reference to this figure?} against general formal interfaces,
  enabling mixing and matching of different realizations of each interface,
  without doing any new proofs that peek beneath abstraction boundaries.

@@

\paragraph{Paper Organization:} In Section~\ref{sec:lts} we introduce our
flavor of the labeled transition systems formalism, including a definition of
trace refinement. In Section~\ref{sec:store-atomicity}, we show a generic
decomposition of any multi-processor system independently of the memory model
that it implements, and we discuss the store-atomicity property of the memory
sub-component. In Section~\ref{sec:sc} we
give a simple formal model of sequential consistency.  The
following sections refine the two main subcomponents from Figure~\ref{zoom}.
Section~\ref{sec:ooo} discusses definition and verification of a speculative
processor model, and Sections~\ref{sec:cc} and~\ref{sec:ccproof} respectively
define and prove our hierarchical cache-coherence protocol.  Finally, in
Section~\ref{sec:finalresult} we show the whole-system modular proof of our
complex system, finishing with discussion of related work and other conclusions in
Sections~\ref{relatedWork} and \ref{sec:conclusion}.
